How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?

### 2024-02-13 outputs/paper/paper.qmd
new table, new references, lots of other changes

ghs_data <- read_excel(here::here("inputs/data/GHS_index.xlsx"))
world_map <- st_read(here::here("inputs/data/world/World_Countries__Generalized_.shp"))
life_exp <- read.csv(here::here("inputs/data/life_table.csv"), stringsAsFactors = FALSE)
vote_share <- read.csv(here::here("inputs/data/Vote_share_2020_data.csv"), stringsAsFactors = FALSE)
deaths <- read.csv(here::here("inputs/data/covid-19-death-rates.csv"), stringsAsFactors = FALSE)

@@

@@

From all the data provided from @citearticle, @citedata1, and @citedata2, we mainly utilize the @data3, @data8, @data12, and @citedata2. The data from @citearticle are quite organized and do not require excessive cleaning for us to proceed.

Since we aim to analyze people's political preferences on a state level and the data we collected from @data12 is on a county level, we create a new data frame "vote_data_by_state" to summarize all the votes by state. We utilize the "group_by" function to form new variables on a state level. We calculated each the sum of the total votes, total Republican party votes, total Democratic party votes, and total vote difference between the two parties by county and created new variables that calculate the sum of each of these. We also create variables on the average percentage of votes each party receives and the difference by each state.

To find the total number of votes in the Top 10 states with the highest proportion of Republican votes, we create another new data frame "sorted_states" from the "vote_data_by_state" data frame. We include a new variable "proportion_republican" that calculates the proportion of votes the Republican party receives from each state and orders the states by the highest proportion of Republican votes. 
The data from @data2 does not have column names for each column. To retrieve data by column we assigned column names "State" and "Death_Rate" to the variables of interest. And then, to find the top 10 states with the highest death rates, we ranked the states based on death rate. 
Lastly, to create the final figure that incorporates the COVID death rate and people's political preference by state, we merged the two data from @data12 and @citedata2. And subsequently did our work from there.

@@

Our results are summarized in the following figures. @fig-ELE illustrates the trend of life expectancy at birth across different racial groups over time. @tbl-life-exp provides a more detailed breakdown of the life expectancy before and during COVID for Asian, American Indian, Alaska Native and other racial communities.

@fig-GHS also allows us to visualize the Global Health Security (GHS) Index which is designed to assess a country's capabilities to prevent, detect, and respond to significant infectious disease outbreaks. A higher Global Health Security (GHS) Index score indicates a stronger health system capacity for pandemic readiness. The GHS Index assesses the readiness of countries for pandemics and other significant biological threat emergencies, using a comprehensive framework that includes measurements grouped into six broad categories: prevention, detection, response, health system capacity, compliance with international norms, and risk environment. Higher scores suggest that a country has a more robust health system capacity, with better capabilities to prevent, detect, respond to, and mitigate the spread of an epidemic.

Unfortunately, due to the unavailability of data over time for Asian, American Indian, and Alaska Native communities, they have been excluded from the time series graph.

An intriguing observation is the consistently higher life expectancy among Hispanic individuals compared to other groups, even amidst the challenges posed by COVID-19. On the other hand, Black individuals l have consistently exhibited lower life expectancy, which further declined notably in 2021, reaching just over 70.8 years old. The life expectancy trends of white people and all other races and origins remain close together throughout the 14 years, with minimum variance. Specifically, the life expectancy for White individuals decreased from 78.8 to 76.4 years old from pre-pandemic levels in 2019 to 2021, while for Hispanic individuals, it dropped by 4.2 years, and for Black individuals, it declined by 4 years during the same period.

@@

With a chart that depicts the change in life expectancy each year by ethnicity, we can better gain a more comprehensive understanding of the impact of COVID by ethnicity. While the declines in life expectancy are smaller in magnitude from 2020 to 2021, they are notably minimal for Asians and Hispanics, with reductions of -0.1 and -0.2 years respectively. (Insert some sources tmrw)

Numerous sources have indicated a correlation between a state's political affiliation and its handling of COVID issues (sources to be inserted tomorrow). To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in Figure @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths that rank the top ten states with the highest death rates from COVID-19 per 100,000 people to examine the potential correlation between party preferences and COVID-related deaths. Notably, We found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. 
(Insert quotes about the republican party and COVID tmrw)

>>>>

ghs_data <- read_excel("C:/Users/35911/Documents/US-Covid-Analysis/inputs/data/GHS_index.xlsx")
world_map <- st_read("C:/Users/35911/Documents/US-Covid-Analysis/inputs/data/world/World_Countries__Generalized_.shp")
life_exp <- read.csv("C:/Users/35911/Documents/US-Covid-Analysis/inputs/data/life_table.csv", stringsAsFactors = FALSE)
vote_share <- read.csv("C:/Users/35911/Documents/US-Covid-Analysis/inputs/data/Vote_share_2020_data.csv", stringsAsFactors = FALSE)
deaths <- read.csv("C:/Users/35911/Documents/US-Covid-Analysis/inputs/data/covid-19-death-rates.csv", stringsAsFactors = FALSE)

# ghs_data <- read_excel(here::here("inputs/data/GHS_index.xlsx"))
# world_map <- st_read(here::here("inputs/data/world/World_Countries__Generalized_.shp"))
# life_exp <- read.csv(here::here("inputs/data/life_table.csv"), stringsAsFactors = FALSE)
# vote_share <- read.csv(here::here("inputs/data/Vote_share_2020_data.csv"), stringsAsFactors = FALSE)
# deaths <- read.csv(here::here("inputs/data/covid-19-death-rates.csv"), stringsAsFactors = FALSE)

@@

@fig-GHS allows us to visualize the Global Health Security (GHS) Index which is designed to assess a country's capabilities to prevent, detect, and respond to significant infectious disease outbreaks. A higher Global Health Security (GHS) Index score indicates a stronger health system capacity for pandemic readiness. The GHS Index assesses the readiness of countries for pandemics and other significant biological threat emergencies, using a comprehensive framework that includes measurements grouped into six broad categories: prevention, detection, response, health system capacity, compliance with international norms, and risk environment. Higher scores suggest that a country has a more robust health system capacity, with better capabilities to prevent, detect, respond to, and mitigate the spread of an epidemic.

@@

From all the data provided from @citearticle, @citedata1, and @citedata2, we mainly utilize  @data3, @data8, @data12, and @citedata2. The data from @citearticle is well-organized and require minimal cleaning for analysis.

The data collected from @data12 is at the county level as depicted on @tbl-sample. However, since our analysis focuses on people's political preferences at the state level, we create a new data frame called "vote_data_by_state" to aggregate voting data by state. We utilize the "group_by" function to generate new variables at the state level, calculating the total votes, total Republican party votes, total Democratic party votes, and the total vote difference between the two parties by county. Additionally, We also create variables for the average percentage of votes each party receives and the difference by each state.

To determine the total number of votes in the Top 10 states with the highest proportion of Republican votes, we create another new data frame called "sorted_states" from the "vote_data_by_state" data frame. We include a new variable "proportion_republican" that calculates the proportion of votes the Republican party receives from each state and orders the states by the highest proportion of Republican votes.

```{r}
#| echo: false
#| eval: true
#| label: tbl-sample
#| tbl-cap: Sample of Vote Share Data
#| warning: false

# Take a sample of the data (let's say the first 5 rows)
vote_share_sample <- head(vote_share, 5)

# Print the sample table
kable(vote_share_sample, 
      format = "markdown", 
      col.names = c("State Name", "County FIPS", "County Name", "Votes GOP", "Votes DEM", "Total Votes", "Difference", "Percentage GOP", "Percentage DEM", "Percentage Point Difference")
)

The data from @data2 lacks column names for each column. To retrieve data by column, we assign column names "State" and "Death_Rate" to the variables of interest. Subsequently, to find the top 10 states with the highest death rates, we rank the states based on death rate. Lastly, to create the final figure that incorporates the COVID death rate and people's political preferences by state, we merge the two datasets from @data12 and @citedata2 and proceed with our analysis.

@@

Our results are summarized in the following figures. @fig-ELE illustrates the trend of life expectancy at birth across different racial groups over time. Unfortunately, due to the unavailability of data over time for Asian, American Indian, and Alaska Native communities, they have been excluded from the time series graph. @tbl-life-exp provides a more detailed breakdown of the life expectancy before and during COVID for Asian, American Indian, Alaska Native and other racial communities.

An intriguing observation from @fig-ELE is the consistently higher life expectancy among Hispanic individuals compared to other groups, even amidst the challenges posed by COVID-19. On the other hand, Black individuals have consistently exhibited lower life expectancy, which further declined notably in 2021, reaching just over 70.8 years old. The life expectancy trends of white people and all other races and origins remain close together throughout the 14 years, with minimum variance.

While all racial groups experienced declines in their average life expectancy, the declines vary greatly. The life expectancy for White individuals decreased from 78.8 to 76.4 years old from pre-pandemic levels in 2019 to 2021, while for Hispanic individuals, it dropped by 4.2 years, and for Black individuals, it declined by 4 years during the same period. Our data aligns with the findings of @andrasfay_goldman_2022, which states that the Black and Hispanic populations experienced twice as much life expectancy decrease as the white population."

@@

With a chart that depicts the change in life expectancy each year by ethnicity, we can better gain a more comprehensive understanding of the impact of COVID by ethnicity. While the declines in life expectancy are smaller in magnitude from 2020 to 2021, they are notably minimal for Asians and Hispanics, with reductions of -0.1 and -0.2 years respectively.

Various studies have confirmed the existence of correlation between a state's political affiliation and its handling of COVID issues (@datz_2022). To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in Figure @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths that rank the top ten states with the highest death rates from COVID-19 per 100,000 people to examine the potential correlation between party preferences and COVID-related deaths. Notably, We found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates observed in Republican-leaning states may be attributed to their preference for less stringent measures compared to Democratic-leaning states, which tend to favor stricter measures (@vandusky-allen_shvetsova_2021).
